Cluster Difference Imaging Photometric Survey. III. An Open Cluster With a 300\,pc Halo

### 2021-02-17 cluster_paper/ms.tex
update paper

% Context, aims, methods, results, conclusions.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations surrounding nearby open clusters. 
  The stars in these halos could have escaped from the cores of the
  cluster, or they could be dispersed remnants of the initial star formation complex.
  They could also be field stars---the false positives of overly
  aggressive clustering algorithms.
  In this study, we focus on a halo with a length of 300\,pc reported
  around the $\approx$120\,Myr open cluster NGC\,2516.
  %TODO: says who?
  The classical tidal radius of the cluster is $\sim$10\,pc.
  Combining photometry from Gaia, rotation periods from TESS, and
  lithium measurements from Gaia-ESO and GALAH, we find that the halo
  of NGC\,2516 is real and coeval with the core.
  Two in three kinematically selected halo members 25-250\,pc from the
  cluster center show rotation periods consistent with a
  gyrochronological age of 120\,Myr.
  A comparison sample of field stars shows no such trend.
  The rotation period and lithium abundances of the cluster stars are
  also anti-correlated at fixed stellar mass, as is the case with
  comparable stellar populations.
  This work expands the set of confirmed NGC\,2516 members by a factor
  of $\approx$2, and quantifies the degree to which Gaia-based
  analyses can identify dispersed stellar populations against a
  background of field star contaminants.
  We highlight the implications for spectroscopic survey targeting,
  open cluster dispersal, and planet searches around young stars.
\end{abstract}

@@

From 10 to 1000\,Myr, the portion of the cluster that survives gas
dispersal suffers an onslaught that typically unbinds the cluster.
Internal to the cluster, mass loss from supernovae and AGB winds, and
close stellar encounters (``relaxation'') are important over 10s and
100s of megayears, respectively (CITE, CITE).  External to the
cluster, collisions with GMCs (Spitzer 1958), and perturbations from
the galactic tide in both the radial (CITE) and vertical (CITE)
dimensions further drive stars toward the ``high velocity tails'' of
the stellar energy distribution.

@@

Identifying the stars that have dispersed from their natal clusters
into the galactic field is of interest for a few reasons.  One 
reason is understanding the predominant conditions under which stars and star
clusters form (e.g., Wright Mamajek 2018 ScoCen), as well as
the processes that drive cluster dissolution (including gas dispersal,
internal and external heating mechanisms, and radial migration across
the galactic disk).
Another is to understand the Sun's birth conditions---
did it form in an open
cluster? If so, how massive was cluster, and is there any hope at
identifying the stars that formed near the Sun (e.g., Adams 2010)?
Finally, the dispersal of open clusters also provides perhaps
the best test case for the concept of ``chemical tagging'', aka.\
``galactic archaeology'' (CITE).  If you can't chemically tag stars
that are kinematically associated, why would you be able to do so once
they've dispersed throughout the galactic disk?

@@

The identification of low-density (GIVE A NUMBER DENSITY) stellar
associations, though possible before Gaia (e.g.,
Zuckerman+ CITE, Bergond+01), has reached a breakthrough pace in the present era.
Some low-density moving groups have no discernable
cores (e.g., Psc Eri CITE, mu Tau CITE), or are hierarchically
sub-structured (e.g., USco CITE, Taurus[?]).
However other
low-density
halos are associated with previously known open clusters (e.g.,
Kounkel+19, Kounkel+20, Meingast+21).  In some
cases, these halos correspond to tidal tails, e.g., in the Hyades
(CITE Meingast+Alves 20, Roser+20), and in the Ursa Major moving
group (CITE Gagne+20). 
However, applying different clustering methods on the
Gaia data yields different results (CITE: Hunt \& Reffert 2020).
Using say Gaussian Mixture Modelling, or any analogous method that
requires ``clusters'' to be ellipses in phase-space unsurprisingly
yields open clusters that are roughly elliptical.  Some unsupervised
clustering methods lead to similar results \citep[][hereafter
\citetalias{cantatgaudin_gaia_2018}]{cantatgaudin_gaia_2018}.  Other
unsupervised methods
have been found to yield dispersed and asymmetric structures
with number densities down to 100 times less than the field, spread
over many hundreds of parsecs
(e.g., \citep[][hereafter
\citetalias{kounkel_untangling_2019}]{kounkel_untangling_2019}; Meingast+21).

@@

As part of a broader project of identifying a large and clean sample of
young stars for a transit search, we focus in this paper on a 
modest question: in just a single rich southern open cluster, is the
cluster halo coeval with the core?  We chose for this
NGC\,2516 for this analysis since it is young ($\sim$120\,Myr; B3V at
the turnoff) and close
($d\approx400$\,pc) enough to facilitate rotation measurements using
TESS, and some degree of spectroscopic analysis.  
Its halo, spanning $\approx$300\,pc, is also quite large.

We want to know: is
the halo real? To what extent can we use Gaia alone to reliably identify
age-dated needles in the haystack of boring field stars?  And more
generally, what are the implications for the evolution of open clusters
if they do have halos?

>>>>

% Context, aims, methods, results, conclusions.
\begin{abstract}
  Recent analyses of the Gaia data have reported the existence of
  diffuse stellar populations surrounding nearby open clusters. 
  The stars in these diffuse halos could have escaped from the cores of the
  cluster, or they could be dispersed remnants of the initial star formation complex.
  They could also be field stars---the false positives of overly
  aggressive clustering algorithms.
  In this study, we focus on a halo with a length of 300\,pc reported
  around the $\approx$120\,Myr open cluster NGC\,2516.
  %TODO: says who?
  The classical tidal radius of the cluster is $\approx$10\,pc.
  Combining photometry from Gaia, rotation periods from TESS, and
  lithium measurements from Gaia-ESO and GALAH, we find that the halo
  of NGC\,2516 is real and coeval with the core.
  Two in three kinematically selected halo members 25-250\,pc from the
  cluster center show rotation periods consistent with a
  gyrochronological age of 120\,Myr.
  A comparison sample of field stars shows no such trend.
  The rotation period and lithium abundances of the cluster stars are
  also anti-correlated at fixed stellar mass, as has been observed in
  similar stellar populations.
  This work expands the set of confirmed NGC\,2516 members by a factor
  of $\approx$2, and quantifies the degree to which Gaia-based
  analyses can identify dispersed stellar populations against a
  background of field star contaminants.
  We highlight the implications for spectroscopic survey targeting,
  open cluster dispersal, and planet searches around young stars.
\end{abstract}

@@

From 10 to 1000\,Myr, the portion of the cluster that survives gas
dispersal suffers an onslaught that typically unbinds the remaining stars.
Internal to the cluster, mass loss from supernovae and AGB winds, and
close stellar encounters (``relaxation'') are important over 10s and
100s of megayears, respectively (CITE, CITE).  External to the
cluster, collisions with GMCs (Spitzer 1958), and perturbations from
the galactic tide in both the radial (CITE) and vertical (CITE)
dimensions further drive stars toward the ``high velocity tails'' of
the stellar energy distribution.

@@

Identifying the stars that have dispersed from their natal clusters
into the galactic field is of interest for a few reasons.  One 
is to understand the predominant conditions under which stars and star
clusters form (e.g., Wright Mamajek 2018 ScoCen), in addition to
the processes that drive cluster dissolution.
Another is to understand the Sun's birth conditions---
did it form in an open
cluster, and if so, what type? Is there any hope at
identifying the stars that formed near the Sun (e.g., Adams 2010)?
Finally, the dispersal of open clusters also provides 
a strong test case for ``chemical tagging'', aka.\
``galactic archaeology'' for stars in the thin and thick disk of
the Galaxy (CITE).  If one cannot chemically tag stars
that are known to be kinematically associated, why would 
it be possible to do so once
they are radially mixed throughout the galactic disk?

@@

The identification of low-density (GIVE A NUMBER DENSITY) stellar
associations, though possible before Gaia (e.g.,
Zuckerman+ CITE, Bergond+01), has reached a breakthrough pace in the present era.
Some low-density moving groups have no discernable
cores (e.g., Psc Eri CITE, mu Tau CITE); others are hierarchically
sub-structured (e.g., USco CITE, Taurus[?]).
The focus of this work is
low-density
halos that are associated with previously known open clusters (e.g.,
Kounkel+19, Kounkel+20, Meingast+21).  In some
cases, these halos correspond to tidal tails, e.g., in the Hyades
(CITE Meingast+Alves 20, Roser+20), and in the Ursa Major moving
group (CITE Gagne+20). 
However, applying different clustering methods on the
Gaia data yields different results (CITE: Hunt \& Reffert 2020).
Using say Gaussian Mixture Modelling, or any analogous method that
requires ``clusters'' to be ellipses in phase-space unsurprisingly
yields open clusters that are roughly elliptical.  Some unsupervised
clustering methods lead to similar results \citep[][hereafter
\citetalias{cantatgaudin_gaia_2018}]{cantatgaudin_gaia_2018}.  Other
unsupervised methods
have yielded dispersed and asymmetric structures
with number densities down to 100 times less than the field, spread
over many hundreds of parsecs
(e.g., \citep[][hereafter
\citetalias{kounkel_untangling_2019}]{kounkel_untangling_2019}; Meingast+21).

@@

As part of a broader project of identifying a large and clean sample of
young stars for a transit search, we focus in this paper on a 
modest question: in just a single rich southern open cluster, is the
cluster halo coeval with the core?  We chose
NGC\,2516 for this analysis since it is young ($\sim$120\,Myr; B3V at
the turnoff) and close
($d\approx400$\,pc) enough to facilitate rotation measurements using
TESS, and some degree of spectroscopic analysis.  
Its halo, spanning $\approx$300\,pc, is also quite large.
We want to know: is
the halo real? To what extent can we use Gaia alone to reliably identify
age-dated needles in the haystack of boring field stars?  And more
generally, what are the implications for the evolution of open clusters
if they do have halos?
